Dialect: A Compact Line Notation Based on SMILES

### 2022-03-18 manuscript.tex
Clarifies introductory discussion of SMILES. refs #85

Despite its many applications, SMILES remains a language with an incomplete specification. Weininger's 1988 paper (referred to herein as "the Paper") leaves the following either fully or incompletely specified: (1) stereochemical configuration; (2) double bond conformation; (3) the meaning handling of "aromatic" features; (4) the computation of implicit hydrogen count, especially as related to aromatic features; (5) formal syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.

Earlier descriptions of SMILES were published in various media, although these are not cited by Weininger. An accessible example is a 1987 report to the US Environmental Protection Agency.\cite{anderson:1987} Although the language described in that report shares many features with later descriptions, there are important differences as well.

Since 1988, a few sources have attempted to clarify or revise Weininger's 1988 description of SMILES. A 2003 book chapter by Weininger (referred to herein as "the Book Chapter") addressed some of the limitations of the paper. The Daylight Theory Manual ("the Manual"), a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc. ("Daylight"), recapitulates much of the material in the Paper and the Book Chapter.\cite{daylightTheory}.

In 2007 a documentation effort that would become known as OpenSMILES began.\cite{openSMILES} OpenSMILES was conceived as "a non-proprietary specification for the SMILES language," and it addressed many of the points left open by previous SMILES documentation efforts. Noteworthy contributions included: the first formal grammar; many refinements around stereochemistry; and introduction of the idea of "standard form." Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count. OpenSMILES also left several points of syntactic and semantic ambiguity unaddressed.

>>>>

Despite its many applications, SMILES remains a language with an incomplete specification. Weininger's 1988 paper leaves the following either fully or incompletely specified: (1) stereochemical configuration; (2) double bond conformation; (3) the meaning handling of "aromatic" features; (4) the computation of implicit hydrogen count, especially as related to aromatic features; (5) formal syntax; (6) constraints on quantities such as mass number and charge; and (7) enumeration of error states.

Earlier descriptions of SMILES were published in various media, although these are not cited by Weininger. An currently-accessible example is a 1987 report to the US Environmental Protection Agency.\cite{anderson:1987} Although the language described there shares many features with later descriptions, there are important differences as well.

Since 1988, two authoritative sources have attempted to clarify or revise Weininger's 1988 description of SMILES. A 2003 book chapter by Weininger addressed some of the limitations of the paper.\cite{weininger:2008} The Daylight Theory Manual, a website maintained by SMILES's corporate sponsor, Daylight Chemical Information Systems, Inc. ("Daylight"), recapitulates previously-published material.\cite{daylightTheory}. The lack of other written sources describing SMILES is consistent with Weininger's reported disinterest in traditional scientific publication.\cite{gasteiger:2018}

In 2007 a documentation effort that would become known as OpenSMILES began.\cite{openSMILES} OpenSMILES was conceived as "a non-proprietary specification for the SMILES language." Many of the points left open by previous SMILES documentation efforts were addressed. Noteworthy contributions included: the first formal grammar, which was later adapted;\cite{denjoed:2010} many refinements around stereochemistry; and the introduction of "standard form." Absent were detailed procedures for assigning and interpreting aromatic features, and a detailed procedure for computing implicit hydrogen count. OpenSMILES also left several points of syntactic and semantic ambiguity unsettled, while introducing extensions of its own.
